---
title: "Exam"
author: "Nuttapong Mekvipad"
date: "6/8/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Questions 0

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ggrepel)
library(DESeq2)
set.seed(2019)
theme_set(theme_bw())
```

## Question 1:

To get bed files containing RefSeq track and CpG island track of hg18, I used Tools > Table browser menu in Genome browser to download both tracks of assembly hg18.

After downloading RefSeq track, the -300 to -1 region around TSSs was selected using below command. The option -l was for specifying the size of region upstream, -r for region downstream and option -s to specify bedtools to select region based on strand.

```{bash eval=FALSE}
bedtools flank -i hg18_refseq.bed -g human.hg18.genome.txt -l 300 -r 0 -s > hg18_refseq_ups300.bed
```

__Question 1.1.  What percentage of promoter regions overlap CpG islands with at least 1 bp?__

To find the percentage of promoter regions overlap CpG islands with at least 1 bp overlapping, I used the bedtools intersect as in below chunk. The hg18_refseq_ups300.bed which contain promoter region was intersected with hg18_cpg.bed, and the potion -v was used to return promoter without any overlapping. We then count the number of line of intersection results.

```{bash eval=FALSE}
bedtools intersect -a hg18_refseq_ups300.bed -b hg18_cpg.bed -v | wc -l # 30873
wc -l hg18_refseq_ups300.bed # 72711
# overlapping promoter = 72711-30873 = 41838
```

In total 57.54 (41838/72711 * 100) percent of promoter regions overlap CpG islands. 

__Question 1.2. What percentage of CpG islands overlap promoter regions with at least 1 bp?__

Similar to question 1.1, I also used bedtools intersect here, but changing input to -a option to file containing CpG island and input to -b option to file containing promoter regions.

```{bash eval=FALSE}
bedtools intersect -a hg18_cpg.bed -b hg18_refseq_ups300.bed -v | wc -l # 15043
wc -l hg18_cpg.bed # 28226
# overlapping CpG = 28226 - 15043 = 13183
```

In total 46.71 percent of promoter regions overlap CpG islands.


